A Distributed Segment-based Approach for Topology-agnostic Nano Networks Routing

### 2013-09-20 PAPER/implementation.tex
Removed some contracted verbal forms, and typos

As discussed above, one of the main design challenge of DNA
Self-Assembled systems is the limited resource available for
implementing both computations and routing decisions for each node.
For this reason, assuming a budget of $10^4$ CNFETs for each network's
node~\cite{liu_jetcs} we estimated the required resource for
implementing the entire DiSR block. An not optimized VHDL description
of the DiSR circuitry has been written and synthesized at gate-level.
Considering the specific layout of each single logic elements ( NAND,
full-adder, latch etc.), it has been possible to get a rough estimate
of the number of transistors necessary for implementing the DiSR
logic.  Figure~\ref{fig:imp_trend} shows the results of synthesis in
terms of number of devices (CNFETs) versus the number of network nodes
while the network scales up from $10\times10$ to $100\times100$ nodes.
\begin{figure}
  \centering
  \includegraphics[width=0.48\textwidth]{pictures/imp_rslt.eps}
  \caption{DiSR synthesis results: N is the number of network's node.}
 \label{fig:imp_trend}
\end{figure}
In particular, observing the Fig.\ref{fig:imp_trend} can 
be pointed out that the proposed implementation occupy  about 20\% of the node budget. 
Further, more than the absolute number of devices itself, it's interesting to
observe that, while the node have a squared increment, the circuitry
complexity which implements the DiSR algorithm increases with a slowly
growing trend.  
An intuitive explaination of this is the relatively simple logic of
DiSR, whose behaviour is almost code on scalable storage structures.
For example, the number of registers implementing the
\emph{link\_visited[]} and \emph{link\_tvisited[]} table follow the
logaritmic function:
\begin{equation}
  \label{eq:imp_trend}
  N_{reg}=2\cdot N_{port} \cdot log_2(N)
\end{equation}
where $N_{port}$ is the number of the router's ports, $N$ is the number 
of network's nodes while the term $2$ takes in account that \emph{link\_visited[]} and 
\emph{link\_tvisited[]} are two separate set of registers. While this number is fixed by 
the number of network's node, the control circuitry could be optimized applying more effort 
in future designs of this building block.

>>>>

As discussed above, one of the main design challenge of DNA
Self-Assembled systems is the limited resource available for
implementing both computations and routing decisions for each node.
For this reason, assuming a budget of $10^4$ CNFETs for each network's
node~\cite{liu_jetcs} we estimated the required resource for
implementing the entire DiSR block. An not optimized VHDL description
of the DiSR circuitry has been written and synthesized at gate-level.
Considering the specific layout of each single logic elements ( NAND,
full-adder, latch etc.), it has been possible to get a rough estimate
of the number of transistors necessary for implementing the DiSR
logic.  Figure~\ref{fig:imp_trend} shows the results of synthesis in
terms of number of devices (CNFETs) versus the number of network nodes
while the network scales up from $10\times10$ to $100\times100$ nodes.
\begin{figure}
  \centering
  \includegraphics[width=0.48\textwidth]{pictures/imp_rslt.eps}
  \caption{DiSR synthesis results: N is the number of network's node.}
 \label{fig:imp_trend}
\end{figure}
In particular, observing the Fig.\ref{fig:imp_trend} can 
be pointed out that the proposed implementation occupy  about 20\% of the node budget. 
Further, more than the absolute number of devices itself, it is interesting to
observe that, while the node have a squared increment, the circuitry
complexity which implements the DiSR algorithm increases with a slowly
growing trend.  
An intuitive explaination of this is the relatively simple logic of
DiSR, whose behaviour is almost code on scalable storage structures.
For example, the number of registers implementing the
\emph{link\_visited[]} and \emph{link\_tvisited[]} table follow the
logaritmic function:
\begin{equation}
  \label{eq:imp_trend}
  N_{reg}=2\cdot N_{port} \cdot log_2(N)
\end{equation}
where $N_{port}$ is the number of the router's ports, $N$ is the number 
of network's nodes while the term $2$ takes in account that \emph{link\_visited[]} and 
\emph{link\_tvisited[]} are two separate set of registers. While this number is fixed by 
the number of network's node, the control circuitry could be optimized applying more effort 
in future designs of this building block.

### 2013-09-20 PAPER/introduction.tex
Removed some contracted verbal forms, and typos

Exploring long term alternatives to the CMOS technology is gaining
more and more revelance as the scaling trend of such devices keeps
introducing new challenges. Power density, defect tolerance, testing
costs and wire delays are only a few of the many critical aspects
involved~\cite{itrs12}. While software parallelism and multicore
approaches~\cite{TODO1} are partially mitigating the impact of such
constraints on performances, it is likely that the growing computing demand will
eventually need even more radical architectural modifications and new
paradigms in order to address the Computer Design challenges of the next
decades.

In the last years, Self-assembled nanoscale architectures~\cite{winfree1998, yan2003}
are emerging as a promising technology due their tera/peta scale of
integration, defect tolerance tolerance and huge potential computing
capabilities. These technologies are certainly still at
their early stage of development, however different laboratory demos and
prof-of-concepts architectures have been presented~\cite{patwardhan2004, patwardhan2006, patwardhan2006_1, pistol2009}.
The main idea behind this approach is to exploit the physical regularity and
stabiliy of DNA structures in order to create a scaffold onto which
nano-devices (e.g. nanowires and nanotubes~\cite{bachtold2001, tans1998, cui2001}) can be
attached. This can be achieved by designing approapriate complementary DNA tags for
each terminal to be placed, so that a nano device will be attached
only where its own DNA tag matches a complementary tag on the DNA grid
scaffold (see figure~\ref{fig:dna_tag}.
\begin{figure}
  \centering
    \includegraphics[width=0.40\textwidth]{pictures/dna2b.eps}
  \caption{DNA tag matching}
  \label{fig:dna_tag}
\end{figure}
A detailed description of the chemical properties involved is far
beyond the scope of this paper (see also~\cite{braun1998, seeman1999}), so we will focus on
the three main challenges that this new fabrication process introduces
in Computer Design: (i)limited node complexity, (ii) large scale
randomness and (iii) high defect rates.

The \emph{limited node complexity} aspect is directly
related to the use of complementary DNA tags in order to place circuit
components. In a traditional CMOS process the complexity is introduced
using a photolithography mask, so that larger (more complex) circuits
only require larger masks; conversely, self-assembly achieves
complexity by increasing the number of unique DNA tags, since more
different tags means having more control on component placement. Ideally, by specifying a single
and unique tag for each nano transistor terminal, we could exactly
choose where each component would be placed. But the number of DNA
symbols forming the DNA sequence is limited (sequence of 4 nucleobases
G,A,T,C) and so creating many different tags (of a given lenght) would
mean make them more similar to each other, increasing the probability
of incorrect/partial matching. To avoid this problem we should limit
the number of unique tags, thus limiting the complexity at each node.

\emph{Large scale randomness} and \emph{high defect rates} are the other two fundaments
conditions of self-assembled technology: if, as seen, some complexity
can be introduced by designing tag sequences in the DNA grid of each
node, we have no control of where each of these grids will be placed
on the whole network. Further, as a consequence, we cannot guarantee
other typical properties of regular networks, like nodes being
connected to 4 neighbors, having a determined orientation and so on.

These aspects of a DNA-self assembly process lead to some important
architectural implications to be addressed when approaching to the
Computer Design: computational model must be founded on a distributed
network of small computing and storage nodes, randomly placed and
interconnected (see figure~\ref{fig:nana}).  Proof-of-concept of such kind of computational model
can be found in~\cite{patwardhan2006_1}, where instructions and data operands travel
on packets and routed using a topology-agnostic strategy that avoids
deadlock, since no particular regularity can be assumed in such
network.

\begin{figure}
  \centering
    \includegraphics[width=0.40\textwidth]{pictures/dna1.eps}
  \caption{Irregular architecture of computing and storage elements}
  \label{fig:nana}
\end{figure}

In this paper we introduce DiSR, a first effort for a Distributed
Segment-based Routing in large scale DNA self assembled networks.
Given the irregular/high-defect rates scenario assumed as background, is essential
avoiding non-scalable or resource hungry solutions such as
centralized off-chip algorithms, tree-based approaches, virtual channels
or expensive implementations based on hardware redundancy.
Altought we will assume an architecture similar to the one depicted
in~\ref{fig:nana} as background, it
is important to underline that the proposed approach to topology
discovery is more general and not stricly dependent on any of these
architectures.

@@

Several works~\cite{sancho2002, skeie2002, skeie2004, koibuchi2003} have been presented addressing the problem of topology
agnostic routing. While they show interesting
performances, the requirement for virtual channels resources 
to avoid deadlock makes them not suitable for the specific DNA self
assembled networks we are assuming as background.
Other topology independent algorithms~\cite{schroeder1991, koibuchi2001, cherkasova1996} don't use virtual channels to
achieve deadlock freedom, but turn proibitions. 
In particular, up*/down* exploits the creation of a spanning tree of the
topology, then placing bidirectional restrictions by avoiding a packet
to traverse the same link in both up and down directions.
While the hierarchical nature of this approach can lead to uneven traffic
distribution, with many packets traversing upper links (near to the
root), this is quite acceptable in classical wide area networks
topologies with a limited number of nodes. Other approaches such as
FX~\cite{sancho2000} mitigate this
issue, but the set of turn restrictions is still prefixed,
striclty depending on the particular tree root selected. 
Other solutions try to approach the issue of irregular
topologies by limiting the number~\cite{dally1994, duato1997, gomez2004, koibuchi2008} or the
location~\cite{zhang2008, sui2000, flich2008, liu2011} of missing links. This
restriction is clearly unacceptable given the hight-defect rates of
DNA self-assembled networks scenario. For the same reason, we also
avoid considering solutions based on hardware-redundancy to
dynamically recover defects as in~\cite{constantinides2006, kohler2010, kim2006, park2006}.

>>>>

% TODO: re-enable on long paper
%Exploring long term alternatives to the CMOS technology is gaining
%more and more revelance as the scaling trend of such devices keeps
%introducing new challenges. Power density, defect tolerance, testing
%costs and wire delays are only a few of the many critical aspects
%involved~\cite{itrs12}. While software parallelism and multicore
%approaches~\cite{horowitz2004, mudge2001, powell2009} are partially mitigating the impact of such
%constraints on performances, it is likely that the growing computing demand will
%eventually need even more radical architectural modifications and new
%paradigms in order to address the Computer Design challenges of the next
%decades.

In the last years, Self-assembled nanoscale architectures~\cite{winfree1998, yan2003}
are emerging as a promising technology due their tera/peta scale of
integration, defect tolerance tolerance and huge potential computing
capabilities. These technologies are certainly still at
their early stage of development, however different laboratory demos and
prof-of-concepts architectures have been presented~\cite{patwardhan2004, patwardhan2006, patwardhan2006_1, pistol2009}.
The main idea behind this approach is to exploit the physical regularity and
stabiliy of DNA structures in order to create a scaffold onto which
nano-devices (e.g. nanowires and nanotubes~\cite{bachtold2001, tans1998, cui2001}) can be
attached. This can be achieved by designing approapriate complementary DNA tags for
each terminal to be placed, so that a nano device will be attached
only where its own DNA tag matches a complementary tag on the DNA grid
scaffold (see figure~\ref{fig:dna_tag}.

\begin{figure}
  \centering
    \includegraphics[width=0.40\textwidth]{pictures/dna2b.eps}
  \caption{DNA tag matching}
  \label{fig:dna_tag}
\end{figure}

A detailed description of the chemical properties involved is far
beyond the scope of this paper (see also~\cite{braun1998, seeman1999}), so we will focus on
the three main challenges that this new fabrication process introduces
in Computer Design: (i)limited node complexity, (ii) large scale
randomness and (iii) high defect rates.

The \emph{limited node complexity} aspect is directly
related to the use of complementary DNA tags in order to place circuit
components. In a traditional CMOS process the complexity is introduced
using a photolithography mask, so that larger (more complex) circuits
only require larger masks; conversely, self-assembly achieves
complexity by increasing the number of unique DNA tags, since more
different tags means having more control on component placement. Ideally, by specifying a single
and unique tag for each nano transistor terminal, we could exactly
choose where each component would be placed. But the number of DNA
symbols forming the DNA sequence is limited (sequence of 4 nucleobases
G,A,T,C) and so creating many different tags (of a given lenght) would
mean make them more similar to each other, increasing the probability
of incorrect/partial matching. To avoid this problem we should limit
the number of unique tags, thus limiting the complexity at each node.
In particular, a budget of about $10,000$ transistor per node has been
estimated in~\cite{liu_jetcs}.

\emph{Large scale randomness} is the other fundamental condition of
self-assembled technology: DNA tags allow a control placement inside
node grid, but there is no control of where each of these  grids will
be placed on the whole network. As a consequence, other typical
properties of regular networks cannot be guaranteed, e.g. being
connected to a fixed number of neighbors, having a determined
orientation and so on.

Finally, \emph{defect rates}: while they are very hardly tolerated in mask
based top down design , the same nature of a bottom up self
assembled process cannot assume such a deterministic device placement
process.

These aspects of a DNA-self assembly process lead to some important
implications to be addressed when approaching to the design of
computer architecture: a computational model should be founded on a
distributed network of small computing and storage nodes, randomly
placed and interconnected (see figure~\ref{fig:nana}).
Proof-of-concept of such kind of computational model can be found for
example in~\cite{patwardhan2006_1}.  Since no regularity can be
assumed in such network, a topology agnostic strategy that avoids
deadlock should be adopted in order to route packets (e.g.
instructions and data operands) among nodes.  It is important to
underline that, altought we can assume an architecture similar to the
one depicted in~\ref{fig:nana} as background, the topology discovery
problem is more general and not stricly dependent on any of the
underlying computing architectures.

%where instructions and data operands travel
%on packets and routed using a topology-agnostic strategy that avoids
%deadlock, 
\begin{figure}
  \centering
    \includegraphics[width=0.20\textwidth]{pictures/dna1.eps}
  \caption{Irregular topology of computing and storage elements}
  \label{fig:nana}
\end{figure}

The contribution presented in this paper is to bring the advantages of
a segment-based approach to deadlock freedom without requiring any topology graph,
external defect map or centralized algorithm execution.  We introduce DiSR, a first effort for a Distributed
Segment-based Routing in large scale DNA self assembled networks.
While maintaining similar properties to segment-based , the DiSR
approach is not intended to discover the "optimal" segment choice
(ideally reachable with the knowledge of the topology graph) but just
to demonstrate a concrete model that can fit into such complex,
irregular and large sized topologies.
Given the irregular/high-defect rates scenario discussed above, is essential
avoiding non-scalable or resource hungry solutions such as
centralized off-chip algorithms, tree-based approaches, virtual channels
or expensive implementations based on hardware redundancy.

@@

Several works~\cite{sancho2002, skeie2002, skeie2004, koibuchi2003} have been presented addressing the problem of topology
agnostic routing. While they show interesting
performances, the requirement for virtual channels resources 
to avoid deadlock makes them not suitable for the specific DNA self
assembled networks we are assuming as background.
Other topology independent algorithms~\cite{schroeder1991, koibuchi2001, cherkasova1996} do not use virtual channels to
achieve deadlock freedom, but turn proibitions. 
In particular, up*/down* exploits the creation of a spanning tree of the
topology, then placing bidirectional restrictions by avoiding a packet
to traverse the same link in both up and down directions.
While the hierarchical nature of this approach can lead to uneven traffic
distribution, with many packets traversing upper links (near to the
root), this is quite acceptable in classical wide area networks
topologies with a limited number of nodes. Other approaches such as
FX~\cite{sancho2000} mitigate this
issue, but the set of turn restrictions is still prefixed,
striclty depending on the particular tree root selected. 
Other solutions try to approach the issue of irregular
topologies by limiting the number~\cite{dally1994, duato1997, gomez2004, koibuchi2008} or the
location~\cite{zhang2008, sui2000, flich2008, liu2011} of missing links. This
restriction is clearly unacceptable given the hight-defect rates of
DNA self-assembled networks scenario. For the same reason, we also
avoid considering solutions based on hardware-redundancy to
dynamically recover defects as in~\cite{constantinides2006, kohler2010, kim2006, park2006}.
